---
title: "PRISMA installation"
author: "Panagiotis Moulos"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{PRISMA installation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# PRISMA installation

Apart from CRAN and Bioconductor dependencies, PRISMA also depends on:

* The presence of certain system packages
* The presence of the [lassosum](https://github.com/tshmak/lassosum) which is
not on CRAN or Bioconductor
* The presence of certain 3rd party tools which must be findable by PRISMA (can
be installed also with PRISMA functions)

The following outline the whole process.

## Prerequisites

### System packages

Some or all of these may be already installed on the system. There might be
additional ones depending on the Linux distribution prisma is being installed
to. This list will be maintained and updated.

```
sudo apt install -y apt-transport-https software-properties-common \
    build-essential zlib1g-dev libdb-dev libcurl4-openssl-dev libssl-dev \
    libxml2-dev apache2 libfontconfig1-dev libjpeg-dev
```

### R packages

#### Basic installation of Bioconductor

```
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.15")
```

#### Additional packages

The following packages are either on CRAN or on Bioconductor:

```
pkgs <- c("BSgenome.Hsapiens.UCSC.hg19","BSgenome.Hsapiens.UCSC.hg38",
    "Biobase","biomaRt","BSgenome","data.table","GenomicRanges","ggplot2",
    "gwasrapidd","harmonicmeanp","islasso","jsonlite","kableExtra","knitr",
    "liftOver","Matrix","magrittr","methods","openxlsx","pander","parallel",
    "PhenotypeSimulator","quincunx","R.utils","rmarkdown","rmdformats",
    "RcppArmadillo","RMTstat","rrBLUP","rrcov","rsnps","rtracklayer",
    "S4Vectors","scales","scrime","snpStats","SNPRelate",
    "SNPlocs.Hsapiens.dbSNP151.GRCh38","statgenGWAS","SummarizedExperiment",
    "survcomp","testthat","tseries","utils")

BiocManager::install(pkgs)
```

Also, `lassosum` is not on CRAN:

```
library(devtools)
install_github("tshmak/lassosum")
```

### Other external tools and resources

The following tools should be findable by R, for example they should be included
in the system's `PATH` environmental variable.

#### Summary statistics

* [SNPTEST](https://www.well.ox.ac.uk/~gav/snptest/)
* [PLINK](https://www.cog-genomics.org/plink/)

#### PRS extraction

* [PRSice-2](https://www.prsice.info/)

#### Other tools (format conversion and imputation)

* [GTOOL](https://www.well.ox.ac.uk/~cfreeman/software/gwas/gtool.html)
* [QCTOOL](https://www.well.ox.ac.uk/~gav/qctool_v2/)
* [IMPUTE2](http://mathgen.stats.ox.ac.uk/impute/impute_v2.html)
* [IMPUTE2 reference panel files](https://mathgen.stats.ox.ac.uk/impute/1000GP_Phase3.tgz)

## Installation of PRISMA

PRISMA can be installed from GitHub for the time being:

```
library(devtools)
install_github("pmoulos/prisma")
```

# Installation of 3rd party tools

In order to exploit all PRISMA functionalities, certain 3rd party tools for
calculation of summary statistics, PRS extraction, file format conversion and
imputation must be present on the system. Particulalry, 
[PRSice-2](https://www.prsice.info/) must be present and findable, otherwise
PRS extraction functionalities (and also the main purpose of the package) will
be very limited. Furthermore, if you wish to manually run the full test-suite of
the package, certain tests will be skipped if the required 3rd party tools are
not present.

The 3rd party tools can be installed outside R and PRISMA manually or from
withing R via PRISMA package funcitonalities. In the first case, more manual
steps are required but there is the freedom for the user to install these tools
at will and whatever location, as long as this is placed on the system's `PATH`
variable. In the second case, the tools are downloaded and installed 
automatically but they must live in a specific location defined by R faclities
in order to be findable by PRISMA.

## Installation of 3rd party tools outside PRISMA

The installation location can be where the user wishes to. Here, we assume
that the tools are installed in `~/prs-tools`:

```
TOOLS_HOME=~/prs-tools
mkdir -p $TOOLS_HOME
```

### Installation of tools for summary statistics

#### SNPTEST

We assume version 2.5.4 for Ubuntu:

```
mkdir -p $TOOLS_HOME/snptest-2.5.4 && cd $TOOLS_HOME/snptest-2.5.4
wget --no-check-certificate  http://www.well.ox.ac.uk/~gav/resources/snptest_v2.5.4-beta3_linux_x86_64_static.tgz
tar -xvf snptest_v2.5.4-beta3_linux_x86_64_static.tgz
cd snptest_v2.5.4-beta3_linux_x86_64_static
mv * ../ && cd ../
rm -r snptest_v2.5.4-beta3_linux_x86_64_static*
# PRISMA expects to find an executable named snptest
mv snptest_v2.5.4-beta3 snptest 
```

#### PLINK

We assume version 1.90 for 64-bit Linux:

```
mkdir -p $TOOLS_HOME/plink-1.90 && cd $TOOLS_HOME/plink-1.90
wget https://s3.amazonaws.com/plink1-assets/plink_linux_x86_64_20220402.zip
unzip plink_linux_x86_64_20220402.zip
rm plink_linux_x86_64_20220402.zip
```

### Installation of PRS extraction tools

#### Installation of PRSice-2

We assume version 2.3.5 for 64-bit Linux:

```
mkdir -p $TOOLS_HOME/prsice-2.3.5 && cd $TOOLS_HOME/prsice-2.3.5
wget https://github.com/choishingwan/PRSice/releases/download/2.3.5/PRSice_linux.zip
unzip PRSice_linux.zip
rm PRSice_linux.zip
```

### Installation of additional tools

#### Installation of IMPUTE2

We assume version 2.3.2 for Linux:

```
mkdir -p $TOOLS_HOME/impute-2.3.2 && cd $TOOLS_HOME/impute-2.3.2
wget --no-check-certificate https://mathgen.stats.ox.ac.uk/impute/impute_v2.3.2_x86_64_static.tgz
tar -xvf impute_v2.3.2_x86_64_static.
cd impute_v2.3.2_x86_64_static
mv * ../
cd ..
rm -r impute_v2.3.2_x86_64_static*
```

#### Installation of GTOOL

We assume version 0.7.5 for Linux:

```
mkdir -p $TOOLS_HOME/gtool-0.7.5 && cd $TOOLS_HOME/gtool-0.7.5
wget --no-check-certificate  https://www.well.ox.ac.uk/~cfreeman/software/gwas/gtool_v0.7.5_x86_64.tgz
tar -xvf gtool_v0.7.5_x86_64.tgz
rm gtool_v0.7.5_x86_64.tgz
```

#### Installation of QCTOOL

We assume version 2.2.0 for a later version of Linux (e.g. Ubuntu > 18.04). For
older versions, either download version 2.0.8 or GTOOL only should be enough.
PRISMA uses QCTOOL and GTOOL to convert PED files to the GEN format and on later
Linux systems, for some reason GTOOL fails (they are both older tools!)

```
mkdir -p $TOOLS_HOME/qctool-2.2.0 && cd $TOOLS_HOME/qctool-2.2.0
wget --no-check-certificate https://www.well.ox.ac.uk/~gav/resources/qctool_v2.2.0-CentOS_Linux7.8.2003-x86_64.tgz
tar -xvf qctool_v2.2.0-CentOS_Linux7.8.2003-x86_64.tgz
cd qctool_v2.2.0-CentOS\ Linux7.8.2003-x86_64/
mv * ../
rm -r qctool_v2.2.0-CentOS\ Linux7.8.2003-x86_64/ qctool_v2.2.0-CentOS_Linux7.8.2003-x86_64.tgz
```

### Editing the `PATH` variable

The following should now be added to the end of your `~/.bashrc` file or the
end of your `~/.profile` file or any other shell starting file according to the
Linux operating system version. In Windows, a similar path must be followed,
described for example [here](https://docs.oracle.com/en/database/oracle/machine-learning/oml4r/1.5.1/oread/creating-and-modifying-environment-variables-on-windows.html).

```
PRS_TOOLS=~/prs-tools # or wherever the TOOLS_HOME variable points to
export PATH=$PATH:\
    $TOOLS_HOME/snptest-2.5.4:\
    $TOOLS_HOME/plink-1.90:\
    $TOOLS_HOME/prsice-2.3.5:\
    $TOOLS_HOME/impute-2.3.2:\
    $TOOLS_HOME/gtool-0.7.5:\
    $TOOLS_HOME/qctool-2.2.0:
```

Then you should logout and re-login or:

```
source ~/.basrhc
```

for the changes to take effect.

### Installation of IMPUTE2 reference panel

This could take some time. Also, the path to the reference is not stored
somewhere within PRISMA and should be provided every time the `extendGWAS`
function is used.

```
mkdir -p $TOOLS_HOME/impute_reference && cd $TOOLS_HOME/impute_reference
wget --no-check-certificate https://mathgen.stats.ox.ac.uk/impute/1000GP_Phase3.tgz
tar -xvf 1000GP_Phase3.tgz
rm 1000GP_Phase3.tgz
```

The path to the reference file is then 
`$TOOLS_HOME/impute_reference/1000GP_Phase3`.

## Installation of 3rd party tools within PRISMA

The installation location is specific and is defined by `tools::R_user_dir(d)`.
Then, inside this directory, the aforementioned name convention is followed for
each tool, i.e. `tool_name_in_lowercase/tool.version`. If you choose to install
this way instead of manual installation and placing in the `PATH` variable, the 
tools should not be moved otherwise PRISMA will *not* be able to find them. The
user will be prompted to accept the installation path. The download wrapper
downloads, installs and tests the tools.

```{r inst-0, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE, warning=FALSE}
library(prisma)
```

### Installation of all tools

```{r inst-1, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE, warning=FALSE}
downloadExternalTools()
```

### Installation of tools for summary statistics

#### SNPTEST

We assume version 2.5.4 for Ubuntu:

```{r inst-2, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE, warning=FALSE}
downloadSnptest(ver="v2.5.4")
```

#### PLINK

We assume version 1.90 for 64-bit Linux:

```{r inst-3, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE, warning=FALSE}
downloadPlink(ver="v1.90")
# or just
# downloadPlink()
```

### Installation of PRS extraction tools

#### Installation of PRSice-2

We assume version 2.3.5 for 64-bit Linux:

```{r inst-4, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE, warning=FALSE}
downloadPrsice(ver="v2.3.5")
# or just
# downloadPrsice()
```

### Installation of additional tools and resources

#### Installation of IMPUTE2

We assume version 2.3.2 for Linux:

```{r inst-5, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE, warning=FALSE}
downloadImpute2(ver="v2.3.2")
# or just
# downloadImpute2()
```

#### Installation of GTOOL

We assume version 0.7.5 for Linux:

```{r inst-6, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE, warning=FALSE}
downloadGtool(ver="v0.7.5")
# or just
# downloadGtool()
```

#### Installation of QCTOOL

We assume version 2.2.0 for a later version of Linux (e.g. Ubuntu > 18.04). For
older versions, either download version 2.0.8 or GTOOL only should be enough.
PRISMA uses QCTOOL and GTOOL to convert PED files to the GEN format and on later
Linux systems, for some reason GTOOL fails (they are both older tools!)

```{r inst-7, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE, warning=FALSE}
downloadQctool(ver="v2.2.0")
# or just
# downloadQctool()
```

### Installation of IMPUTE2 reference panel

This could take some time. Also, the path to the reference is not stored
somewhere within PRISMA and should be provided every time the `extendGWAS`
function is used.

```{r inst-8, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE, warning=FALSE}
dir.create("TOOLS_HOME/impute_reference")
download1000GP3("TOOLS_HOME/impute_reference")
```

The path to the reference file is then 
`TOOLS_HOME/impute_reference/1000GP_Phase3`.

# Test the PRISMA installation

Prior to each build and release, we perform extensive tests shipping with the
package. If you wish to run these tests (can take some time) then:

```{r inst-9, eval=FALSE, echo=TRUE, tidy=FALSE, message=TRUE, warning=FALSE}
library(testthat)
test_package("prisma")
```
